Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-03-05 paper/manuscript.tex
First pass through my intented revisions  On branch sboresch  Changes to be committed: 	modified:   manuscript.tex 	modified:   sb.bib

% aux. comments Stefan Boresch -> Stefan Boresch
\newcommand{\sbnote}[1]{%
  {\bfseries{}[SB: }%
  {\textcolor{blue}{#1}}{\bfseries{}]}
  }

@@

>>>>

% aux. comments Stefan Boresch -> Stefan Boresch
\newcommand{\sbnote}[1]{%
  {\bfseries{}[SB: }%
  {\textcolor{blue}{#1}}{\bfseries{}]}
  }
\newcommand{\rref}{{\bfseries[REFs]}{}}

@@

The practical implementation of the approaches to carry out alchemial transformations above leads to additional, often non-trivial differences within each methodology. In the description of DC approaches we already pointed out that the two sets of coordinates can be held on top of each other by constraints \rref or restraints. In the latter case, the restraints may couple the two set of coordinates directly \rref, or the restraints may be defined relative to a common reference frame. Finally, in ATM the two copies of the coordinates are held separate.

Concerning SC approaches, Pearlman’s usage of the term single topology not only meant a chimeric construct as depicted in Fig.~{\bfseries XX}, but specifically implied that the coupling parameter $\lambda$ was applied to the force field parameters. E.g., a bond stretching energy term changing from C-H to C-C has the functional form $U_{bond}(\lambda)=K(\lambda)(r-r_0(\lambda))^2$. For $\lambda=0$, $K(\lambda)$ and $r_0(\lambda)$ correspond to the values appropriate for an aliphatic C-H bond, and for $\lambda=1$ to those for an aliphatic C-C bond. The other energy terms are treated analogously. The single topology paradigm as just summarized was the default mode of alchemical transformations in AMBER and GROMOS at that time.  More recently, this implementation of single topology has been referred to as parameter interpolation or mixing \cite{Giese_2018}.

The $\lambda$-dependent interpolation can also be accomplished by mixing the energies and forces of the initial ($\lambda=0$) and final state ($\lambda=1$), e.g., linearly according to $U(\lambda)=(1-\lambda) U_0+\lambda\,U_1$. Other λ-dependencies can be used, and soft-cores \rref may be used for Lennard-Jones and electrostatic interactions to avoid endpoint problems. Giese et al. \cite{Giese_2018} refer to this approach as energy mixing (interpolation).%; it requires two separate calls to the energy/force routines, but since only a fraction of interactions is affected by the alchemical changes, various optimizations are possible to keep the computational overhead manageable.
An early example of energy mixing for SCST is the PERT free energy module of CHARMM (written by Bernie Brooks around 1990 and briefly described in Ref.~\cite{brooks_2009}.

Any SCDT approach requires two energy/force calculations and, thus, uses energy mixing. Therefore, approaches that implement energy mixing together with an alchemical topology (a single set of coordinates) often support both SCST and SCDT. E.g., CHARMM's PERT module just mentioned could be used to set up either of the SC approaches sketched in Fig.~{\bf XX}. Finally, all DC approaches require two energy/force evaluations.

Any energy mixing approach, whether in a SC or DC context, requires two energy/force evaluations. While various optimizations keep the computational overhead significantly below a factor of two,  parameter mixing, if applicable, is the compuationally most efficient approach. However, it is closely tied to SCST and the more complicated the alchemical transformation, the more challenging a strict SCST approach becomes. Also, the mixing of partial charges and the calculation of $\partial U/\partial\lambda$ for the reciprocal space particle-mesh Ewald (PME) summation is tricky. Giese et al.\ described how to do this \cite{Giese_2018}, but most energy mixing implementations calculate the PME term twice and mix the resulting energy and forces.
